---
title: "Separation and non-identification in logistic regression models: a simulation study"
subtitle: "test"
author:
  name: José Reis
affiliation:
  address: "data and code at https://github.com/josemreis/separation_project"
column_numbers: 4
body_textsize: "22pt"
output: 
  posterdown::posterdown_html:
    self_contained: false
bibliography: skeleton.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, cache = TRUE, error = FALSE, fig.align="center", message = FALSE, fig.height = 8, fig.width = 10)
### packages
packs <- c("tidyverse", "kableExtra", "ggridges")
for (pack in packs) {
  
  if (pack %in% installed.packages()){
    
    require(pack, character.only = TRUE)
    
  } else {
    
    install.packages(pack)
    require(pack, character.only = TRUE)
    
  }
}
## turn off scientific notation
options(scipen=999)
## plot settings: default
theme_set(theme_minimal(base_size = 16))
```

# Logistic regression

Logistic regression is a generalized linear model for binary data. It models the conditional distribution of $Y$, a random vector composed of dichotomous observations $(y_1, ...,y_n) \in \{0,1\}$, given $X$, a vector of co-variate variables $(x_i, ...,x_n)$, as a Bernoulli distribution, or $Binomial(1, \pi)$.

$$
y_i|x_i \sim \text{Binomial}(1, \pi_i) 
$$
Where $\pi_i$ is assumed to be a function of a linear component, $\alpha + \beta X_i$, which is modeled using a logit link function.


$$
g(\pi_i) = logit \: \pi_i = log \: \frac{\pi_i}{1 - \pi_i} = \alpha + \beta x \\
\text{where} \\
Pr(Y=1|X) = logit^{-1} \pi_i = \frac{e^{\alpha + \beta x}}{1 + e^{\alpha + \beta x}}
$$

To estimate $\beta_i$ researchers find the coefficient vector which maximizes log-likelihood function.



$$
l(\beta|x_1, ..., x_n) = \sum_{i=1}^n y_i \: log \: \pi_i + (1-y_i) \: log \: (1-\pi_i)
$$


In most cases, the MLE must be approximated using some numerical method (e.g. Newton-Raphson algorithm).


# Separation

**Complete separation** occurs if there is a vector of parameter values $\beta = (\beta_1, ..., \beta_n)$ which partitions the explanatory value at $T$ in such a way that in one side the dependent variable is always $y_i = 1$ and on the other it is always $y_i = 0$ (see @albert1984existence). That is, the linear component can completely discriminate $y$.

$$
\beta x_i > T \: \text{whenever} \: y_i = 1,\\
\beta x_i < T \: \text{whenever} \: y_i = 0
$$


**Quasi complete separation** occurs whenever there is at least one vector of parameter values $\beta = (\beta_1, ..., \beta_n)$ which partitions the explanatory value at $T$ in such a way that the following equality holds for at least one $y_i$ (*idem*).

$$
\beta x_i \geq  T \: \text{whenever} \: y_i = 1 \\ 
\beta x_i \leq  T \: \text{whenever} \: y_i = 0 
$$





```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", fig.cap="Perfect and quasi-perfect separation"}
### function for generating logistic regression data
logistic_sim <- function(n = 100, beta = 0.5, seed = NULL){
  
  ####################################################################################################################
  ### simulate population data for a logistic regression y|x ~ logit^{-1} \pi(x)
  # model details:
      # y|x ~ Binomial(1, \pi) \\
      # \pi = logit^{-1} \pi = \frac{exp(\alpha + \beta X)}{1 + exp(\alpha + \beta X)} \\
      # X ~ N(0,1) \\
      # \alpha = 0
  # n: int; size of the population data to simulate. 
  # beta: numerical; true beta parameter value used to generate the population data, i.e. \pi for y ~ Binomial(1, \pi)
  # seed: int; random seed user defined
  #####################################################################################################################
  
    ### simulate population data
  if (length(seed) == 0) {
    
    random_seed <- sample(.Random.seed, 1) 
    
  } else {
    
    random_seed <- seed
    
  }
  set.seed(seed)
  alpha <- 0 # centered data
  df <- tibble(x = rnorm(n, 0, 1)) %>%
    mutate(pi = plogis((alpha + beta * x)), 
           y = rbinom(n, 1, prob = pi),
           random_seed = random_seed)
  
  return(df)
}


make_separation_plot <- function(n = 100, beta = 0.5, seed = 1234) {
  
  df <- logistic_sim(n = n, beta = beta, seed = seed)
  
  p <- df %>%
    ggplot(aes(x = x, y = y)) + 
    geom_point(size = 4) +
    annotate("rect", xmin = -Inf, xmax = mean(df$x), ymin = -Inf, ymax = .5, fill = "palegreen", alpha = 0.5) +
    annotate("rect", xmin = mean(df$x), xmax = Inf, ymin = -Inf, ymax = .5, fill = "#F8766D", alpha = 0.5)  +
    annotate("rect", xmin = -Inf, xmax = mean(df$x), ymin = Inf, ymax = .5, fill = "#F8766D", alpha = 0.5) +
    annotate("rect", xmin = mean(df$x), xmax = Inf, ymin = Inf, ymax = .5, fill = "palegreen", alpha = 0.5) +
    stat_smooth(method = "glm", 
                method.args = list(family = "binomial"),
                se = FALSE) +
    labs(y = bquote(logit^-1 ~pi[(x)]))
  
  return(p) 
}

p1 <- make_separation_plot(n = 25, beta = 5, seed = 123)
p2 <-  make_separation_plot(n = 25, beta = 3, seed = 123)
cowplot::plot_grid(plotlist = list(p1, p2))
```



## The problem

* MLE goes to infinity
* Nevertheless, most statistical software will *converge* before reaching the MLE and report biased log odds ratios and SEs 
* p-values and Wald confidence intervals cannot be relied on for hypothesis testing


```{r, fig.cap="Negative log-likelihoods"}
plot_compare_likelihoods_n <- function (beta, n, up_to = 20, seed){
  
  ## Define the negative log likelihood function
  logl <- function(beta,x,y){
    y <- y
    x <- as.matrix(x)
    loglik <- sum(-y*log(1 + exp(-(x%*%beta))) - (1-y)*log(1 + exp(x%*%beta)))
    ## turn to df
    out <- tibble(loglikelihood = loglik,
             x = x,
             beta = beta,
             y = y) %>%
      mutate(overlap = if_else((x * beta > 0 & y == 1) | (x * beta < 0 & y == 0), 1, 0))
    return(out)
  }
  
  plot_list <- list()
  for (i in 1:length(n)) {
    b <- beta
    s <- n[i]
    set.seed(seed)
    ## generate the data given the beta
    d <- logistic_sim(beta = b, n = s, seed = seed)
    ## compute the log likelihood and make the plit
    df <-  map_df(seq(from = 0.5, to = b + up_to, by = 0.1), ~ logl(x = d$x, y = d$y, beta = .x)) 
    p <- df %>%
      ggplot(aes(beta, loglikelihood)) + 
      geom_line() + 
      geom_vline(xintercept = b, color = "red", linetype = "dashed") +
      scale_x_continuous(breaks = seq(from = 0, to = up_to, by = 5)) + 
      labs(y = "neg. log likelihood",subtitle = paste0(" proportion separated = ", sum(df$overlap)/nrow(df)))
    
    plot_list[[i]] <- p
    
  }
  
  return(plot_list)
  
}
cowplot::plot_grid(plotlist = plot_compare_likelihoods_n(beta = 5, n = c(25, 1000), seed = 123))

```


# Proposed solutions

## Penalized maximum likelihood 

While MLE estimates do not exist under separation, Firth [-@firth1993bias] showed that by adding the log of the square root of the fisher information matrix as a penalty to the log likelihood function

$$
l_f = l(\beta) + \frac{1}{2} log |I(\beta)|
$$

it becomes is strictly concave and so the penalized log-likelihood will converge on one unique maximum penalized likelihood estimate. 


## Bayesian approaches


In a nutshell, Bayesian approaches focus on estimating the probability distribution of the coefficient given the observed data. This is estimated via the Bayes theorem


$$
\overbrace{Pr(\theta|D)}^\text{Posterior} = \frac{\overbrace{Pr(D|\theta)}^\text{Likelihood} \times\overbrace{Pr(\theta)}^\text{Prior}}{\underbrace{Pr(D)}_\text{Average Likelihood}}
\\
$$
Bayesian methods combine the model of the data (likelihood) with *prior information* about $\beta$. The prior distribution can be used to constrain our likelihood estimates. Researchers have studied how prior distributions of the coefficients may curb separation issues. They have focused on which is the best weakly informed prior for separation issues - i.e. default options.

* Some literature discusses using a $Cauchy(x_0 = 0, \gamma = 2.5)$ distribution [@gelman2008weakly] as default
* Others argue for a $t(df = 7)$ distribution [@ghosh2015use]
* Alternatively, a normal distribution centered around 0 and with an [informed](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations) choice of standard deviation [@rainey2016dealing]


```{r fig.cap="Priors discussed in the literature"}
ggplot(data.frame(x = c(-10, 10)), aes(x)) +
  stat_function(fun = dcauchy, n = 1e3, args = list(location = 0, scale = 2.5), aes(color = "a"), size = 2) +
  stat_function(fun = dt, n = 1e3, args = list(df = 7), aes(color = "b"), size = 2) +
  stat_function(fun = dnorm, n = 1e3, args = list(mean = 0, sd = 2.5), aes(color = "c"), size = 2) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_color_discrete(name = "",
                       labels = c("a" = bquote(beta ~ "~ Cauchy(0, 2.5)"),
                                  "b" = bquote(beta ~ "~ t(7)"),
                                  "c" = bquote(beta ~ "~ Normal(0, 2.5)"))) +
  labs(y = "P(x)") 
```


# Simulation

The robustness of the following methods was put to test in a simulation study.


```{r}
tab <- data.frame(
  "Method" = c(
    "Maximum Likelihood Logistic Model",
    "Maximum Likelihood Probit Model",
    "Maximum Likelihood Linear Probability Model",
    "Penalized Maximum Likelihood Logistic Regression",
    "Bayesian Logistic Regression with $\\beta \\sim Cauchy(0, 2,5) \\: \\alpha \\sim t(0, 0, 10)$",
    "Bayesian Logistic Regression with $\\beta \\sim t(7,0,0) \\: \\alpha \\sim t(0, 0, 10)$",
    "Bayesian Logistic Regression with $\\beta \\sim Normal(0, 2.5) \\: \\alpha \\sim t(0, 0, 10)$"),
  "Point Estimate" = c(
    "MLE",
    "MLE",
    "MLE",
    "PMLE",
    "MAP",
    "MAP",
    "MAP"
  ),
  "CI" = c(
    "Not applicable",
    "Not applicable",
    "Not applicable",
    "Profile Likelihood",
    "Highest Density Probability Interval",
    "Highest Density Probability Interval",
    "Highest Density Probability Interval"
  ),
  "Software" = c(
    "stats::glm() (R 4.0.3)",
    "stats::glm() (R 4.0.3)",
    "stats::lm() (R 4.0.3)",
    "logistif::logistf() (R 4.0.3)",
    "rstanarm::stan_glm() (R 4.0.3)",
    "rstanarm::stan_glm() (R 4.0.3)",
    "rstanarm::stan_glm() (R 4.0.3)"
  ))

tab %>%
  kbl(caption = "Methods Evaluated") %>%
  kable_classic_2(full_width = F, html_font = "Cambria", font_size = 20)

```


Each procedure was evaluated based on:

1. **Bias**, $Bias = E[\hat{\beta}] - \beta$
2. **Mean-squared error**, $MSE = Var(\hat{\beta}) + Bias(\hat{\beta})^2$
3. **Coverage probability of the confidence intervals**, *i.e.* proportion of CIs containing the true parameter



## Data generating process

1. Randomly sample 100 parameters between 2 and 6, $\beta_{sim}$
2. Simulate a population with 1.000.000 observations using the following data generating process

$$
y|x \sim Binomial(1, \pi) \\
\pi = logit^{-1} \pi = \frac{exp(\alpha + \beta_{sim} X)}{1 + exp(\alpha + \beta_{sim} X)} \\
X \sim N(0,1) \\
\alpha = 0 
$$

3.  From each population draw $\frac{1.000.000}{n_{sample}}$ samples of size $n_{sample} = \{25, 50, 75\}$ and check whether perfect separation or quasi-perfect separation occurred [@detect]
4. Keep only samples which contain separation issues. Keep only populations which contain at least 1000 samples with separation issues
5. For each population, and given their true $\beta$, randomly sample 1000 samples to study the estimation procedures

We ended up with 55 simulated populations each with 1000 simulated samples (N = 55000). Only samples of size 25 matched the requirements mentioned above. 


## Results

```{r include=FALSE}
## source the simulation. If the results are already stored, it just loads the data. Simulation takes more than a day.
cur_dir <- getwd()
file <- "../results/simulation/simulation_results.csv"
if (!file.exists(file)){
 setwd("..")
  source("../simulation.R")
  setwd(cur_dir)
   
} else {
  
  data <- read_csv(file)
  
}
```



```{r}
data %>% 
  mutate(estimation_method = if_else(estimation_method == "Bayesian logit model, beta ~ t(1, 0, 2.5)", "Bayesian logit model, beta ~ Cauchy(0, 2.5)", estimation_method)) %>% ## the same but more straightforward
  group_by(estimation_method) %>%
  summarise(`Bias averaged over the 55 populations (#samples = 1000)` = mean(bias),
            `MSE averaged over the 55 populations (#samples = 1000)` = mean(mse),
            `Coverage probability averaged over the 55 populations (#samples = 1000)` = mean(coverage_probability)) %>%
  rename(`estimation method` = estimation_method) %>%
  mutate_if(is.numeric, round, 4) %>%
  arrange(`MSE averaged over the 55 populations (#samples = 1000)`,abs(`Bias averaged over the 55 populations (#samples = 1000)`)) %>%
  kbl(caption = "findings") %>%
  kable_classic_2(full_width = F, html_font = "Cambria", font_size = 20)

```



```{r, fig.cap="Expected value of beta and simulation standard error"}
data %>% # same but more straight forward
  mutate(estimation_method = if_else(str_detect(estimation_method, "1, 0"), "Bayesian logit model, beta ~ Cauchy(0, 2.5)", estimation_method)) %>%
  ggplot(aes(true_beta, ev_beta, color = estimation_method)) + 
  geom_point() + 
  geom_errorbar(aes(ymax = ev_beta + mcse, ymin = ev_beta - mcse), position = "dodge") + 
  facet_wrap(~estimation_method, scales = "free") + 
  theme(strip.text.x = element_blank()) +
  labs(caption = "Error bars are computed using the monte carlo standard error of the 1000 simulations for each of the 55 beta values.",
       color = "estimation method",
       y = "Expected value of the estimated betas",
       x = "True beta") +
  theme(legend.position =  c(0.7, 0.1),
          legend.direction = "vertical")


# data %>% # same but more straight forward
#   mutate(estimation_method = if_else(str_detect(estimation_method, "1, 0"), "Bayesian logit model, beta ~ Cauchy(0, 2.5)", estimation_method)) %>%
#   select(estimation_method, mse, bias, variance = empirical_variance_beta) %>%
#   pivot_longer(cols = c(mse, bias, variance), names_to = "metric", values_to = "value") %>%
#   ggplot()  +
#   ggridges::geom_density_ridges(aes(
#     x = log(value),
#     y = estimation_method,
#     fill = estimation_method),
#     rel_min_height = 0.01,
#     alpha = 0.6) +
#   theme_minimal() +
#   labs(x = "metric logged",
#        y = "estimation method") + 
#   scale_fill_discrete(guide=FALSE) + 
#   facet_wrap(~metric)

```




* With *separated samples* MLE provides very biased estimates of the population parameter
* PML and assigning prior distributions to the parameter curb this problem. Within the realm of MLE methods, the linear probability model seems to be drastically superior than the logistic and probit models.
* Bayesian approaches yielded less biased estimates than PML, except in the case of $\beta \sim Cauchy(0, 2.5)$

# References
